### 2018-04-26 contents/sagit.tex
Updates from ShareLaTeX

\textbf{Results:} At a group level, merged tractography demonstrated that different methods have different fiber distribution characteristics. CST$_{prob}$ is prone to false-positives, and thereby suitable in anatomy with strong priors. CST$_{det}$ and XST are more conservative, but have greater difficulty resolving hemispherical decussation and distant crossing projections. DTT consistently shows the worst reproducibility across the anatomies. Linear regression of rater scores against NOS shows significant (p\textless0.05) correlation of the two sets of scores in filtered tractography. However, correlations are not significant (p\textgreater0.05) for unfiltered tractography.

@@

Diffusion magnetic resonance imaging (dMRI) tractography is an imaging analysis technique that permits non-invasive visualization of white matter anatomy $in$ $vivo$ \cite{Basser2000}.  It is based on the observation that the Brownian motion of water molecules within white matter fibers is constrained by the axonal bundles, and therefore such anisotropic water diffusion can be used to probe tissue microstructure \cite{Bihan2001a}. By scanning in multiple angular directions using diffusion weighted imaging (DWI) sequence, a model of diffusion can be estimated for each voxel of the scanned image volume. 
In the classic single-tensor dMRI (also known as diffusion tensor image; DTI) model \cite{Basser2002}, a single tensor is constructed at each voxel based on Gaussian model of diffusion, which describes the dominant diffusion direction. Using the tensor information, tractography algorithms \cite{Mori2002b} can be used to trace out structures within the DTI volume. The limitation of DTI is that there is insufficient information to resolve areas with crossing fibers with one tensor per voxel. Improvements over the limits of the DTI model, particularly in increasing angular resolution to improve crossing fibers resolution has been a subject of great interest \cite{Fritzsche2010d,Jeurissen2012d,Tuch2002f}.

To improve the crossing fiber information per voxel, significant changes in DWI acquisition strategies are required. These methods include diffusion spectrum imaging (DSI)\cite{Wedeen2005,Wedeen2008} which samples the Q-space in a Cartesian grid, and Q-ball imaging that is based on the Funk-Radon transform \cite{Cho2008,Descoteaux2007a} that samples the Q-space in a spherical shell. These approaches, as well as more complex multi-shell sampling strategies, aim to construct and sharpen an orientation distribution function (ODF) in order to provide better approximate of the underlying diffusion. These methods require complex and long DWI acquisitions that are often unsuitable for clinical applications. Alternatively, methods based on spherical deconvolution (SD) can be used on HARDI scans of lesser angular resolution. The approach is to model the HARDI signal as a convolution of the fiber orientations. The resulting deconvoluted fiber orientation distribution (FOD) shows better discrimination of fiber directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,DellAcqua2007,Tournier2004}. Alternatively, multi-tensor model-based approaches estimate the fiber configuration with the assumption that there are no more than two or three crossings in a given voxel. These include eXtended Streamline Tractography (XST), which can delineate lateral projections of the corticospinal tract in the motor cortex, in DWI scans with 50 gradient directions \cite{Qazi2009}. Other tractography algorithm include stochastically resolving tract propagation based on established white-matter prior probability (probabilistic tractography) \cite{Behrens2007}. Probabilistic approaches however are predominantly image-based, and result in a visitation volume image that needs to be visualized volumetrically.

@@

Finally, we deploy our system to the evaluation of various tractography methods in their ability to delineate neuroanatomy that are highly specific but difficult to image. These include supra and infratentorial structures such as: small white matter bundles in heavy crossing fibers (cranial nerves), pontine decussation (rubrocerebellar pathway) and curved central pathways (fornix, visual and auditory radiations). We recruited 5 neuroanatomists to evaluate and rate the resulting visual reports generated by SAGIT. We then compared the results of NOS scores and the human raters to assess the automated NOS score as a tractography reproducibility metric.

@@

The individual tractography models were also converted to binary spatial images, and the resulting images were transformed to the template space. Multiple binary images were stacked together to form the conjunction percentage overlap image. The overlaps were then visualized using concentric isosurfaces created from step-wise (10\%) thresholds of the underlying overlap volume using the MayaVI data visualization library \cite{Ramachandran2011}. A color lookup table with even visual brightness falloff \cite{VanderWalt2015} was chosen to avoid visual judgment biases from the color presentation. Different viewpoints (axial, coronal, sagittal, and perspective) were created for each visualization, and then composed together to create the visual report.

@@

For the CN VII/VIII, CST$_{prob}$, CST$_{det}$ and XST were all able to delineate the cranial as well as brainstem segments of the nerve (Figure 2). CST$_{prob}$ however also delineated much of the cerebellum. DTT by Slicer was not able to delineate the brainstem portions of the fiber and therefore was scored lower than the other methods. These differences in delineation can be more clearly seen in the merged tracts (Figure 3), where the XST also showed further lateral delineations of CN VII/VIII. In unfiltered tractography, there was much more inter-rater variability for all except CST$_{prob}$. CST$_{det}$ and XST also showed lower average rating when unfiltered (Figure 5). The NOS showed similar trend where DTT resulted in the lowest score (Figure 6). The NOS also showed higher score for CST$_{det}$ when comparing to XST. There were no notable changes in NOS scores in unfiltered tractography.

@@

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{figure2.png}
\caption[Example of the auto-generated visual panels.]{Example of the auto-generated visual panels. The panels that best represent the resulting anatomy are compiled. For anatomical reference, the best representative anatomy image slice is composed with the tract conjunction map. The color scale of the conjunction image represents the percentage reproducible for each region, where 1=100\%. For each algorithm (rows), different visual perspectives are created. Together 6 different anatomies were delineated (columns). To see the full visual reports generated, please see supplemental.}
\label{fig:SGfigure2}
\end{figure}

@@

\afterpage{%
    \clearpage% Flush earlier floats (otherwise order might not be correct)
    \thispagestyle{empty}% empty page style (?)
    \begin{landscape}% Landscape page
\begin{figure}[p]
\includegraphics[width=\linewidth]{figure3.png}
\caption[Example of merged tractography models for each of the techniques and anatomy.]{Example of merged tractography models for each of the techniques and anatomy. The fornix is colored to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data. The other anatomies are coloured by FA intensity. The specific FA measures from each point were sampled from their native DWI space before tractography deformation. The grouping of the similar FA measures across the subjects show that the registrations were accurate. The Red Nuclei shows different metrics (FA and AD) depending on lateralization, in order to highlight the pontine decussation.}
\centering
\label{fig:SGfigure3}
\end{figure}
    \end{landscape}
    \clearpage% Flush page
}

@@

It can be observed that results of averaged tractography image conjunction across a group are visually similar to probabilistic tractography in an individual, and that the group junction image seems to stabilize as the number of subject increase. It can be argued that the conjunction image captures the probable volume across a sampled population for a particular neuroanatomy in the template space. Their exact similarity and differences from probabilistic tractography at a group level should be explored in future studies. Additionally, the fidelity of the merged anatomy is often higher than what would be available from an individual. The result supports the validity of this technical approach. It is possible that the resulting average delineations can be used to generate population-specific anatomy atlases. It's also possible to use fiber density images rather than binary mask for conjunction. However the result of an average fiber density image is difficult to interpret, and therefore should be explored in future studies.

@@

For the specific tractography algorithms, CST$_{prob}$ produces the highest ratings, and performs particularly well in the optical radiation. It also results in more false-positives. This is most evident in CN VII/VIII and CN X results. In these cases the cranial nerve anatomies are very specific and local, and yet CST$_{prob}$ produces more false-positive projections to distant regions. In the fornix there are also erroneous tract extensions into the corpus callosum. Deterministic methods in comparison are more conservative and therefore result in lower scores. They perform well when the anatomy is regional and has finer features. Both CST$_{det}$ and XST performed well with cranial nerves. XST is able to image the rubro-cerebellar pontine decussation with more reliability than CST$_{det}$. While CST$_{prob}$ decussation delineation shows the highest reproducibility, it also produces wider projection coverage that’s harder to interpret. The low DTT score is not surprising, as the single-tensor's inability to resolve crossing-fibers is well known. Based on the result of the study, we recommend the use of CST$_{prob}$ for tasks in which the target anatomy is well defined and false positively can be easily recognized , or when interhemispheric and long projection distances are desired. For exploratory tractography, where filtering locations are not well defined, or when the anatomy features are close together, deterministic methods are recommended. When comparing CST$_{det}$ and XST, it appears that XST is more conservative and thereby results in less false-positives. It also has better ability in resolving hemispherical decussations, as evident by the results of the rubro-cerebellar projections. In practice the two methods are very close in performance. Note however that the comparison is not exhaustive of the possible parameter combinations on either method, and therefore automated tuning of the parameters can answer this question more definitively.

>>>>

\textbf{Results:} At a group level, merged tractography demonstrated that different methods have different fibre distribution characteristics. CST$_{prob}$ is prone to false-positives, and thereby suitable in anatomy with strong priors. CST$_{det}$ and XST are more conservative, but have greater difficulty resolving hemispherical decussation and distant crossing projections. DTT consistently shows the worst reproducibility across the anatomies. Linear regression of rater scores against NOS shows significant (p\textless0.05) correlation of the two sets of scores in filtered tractography. However, correlations are not significant (p\textgreater0.05) for unfiltered tractography.

@@

Diffusion magnetic resonance imaging (dMRI) tractography is an imaging analysis technique that permits non-invasive visualization of white matter anatomy $in$ $vivo$ \cite{Basser2000}.  It is based on the observation that the Brownian motion of water molecules within white matter fibres is constrained by the axonal bundles, and therefore such anisotropic water diffusion can be used to probe tissue microstructure \cite{Bihan2001a}. By scanning in multiple angular directions using diffusion weighted imaging (DWI) sequence, a model of diffusion can be estimated for each voxel of the scanned image volume. 
In the classic single-tensor dMRI (also known as diffusion tensor image; DTI) model \cite{Basser2002}, a single tensor is constructed at each voxel based on Gaussian model of diffusion, which describes the dominant diffusion direction. Using the tensor information, tractography algorithms \cite{Mori2002b} can be used to trace out structures within the DTI volume. The limitation of DTI is that there is insufficient information to resolve areas with crossing fibres with one tensor per voxel. Improvements over the limits of the DTI model, particularly in increasing angular resolution to improve crossing fibres resolution has been a subject of great interest \cite{Fritzsche2010d,Jeurissen2012d,Tuch2002f}.

To improve the crossing fibre information per voxel, significant changes in DWI acquisition strategies are required. These methods include diffusion spectrum imaging (DSI)\cite{Wedeen2005,Wedeen2008} which samples the Q-space in a Cartesian grid, and Q-ball imaging that is based on the Funk-Radon transform \cite{Cho2008,Descoteaux2007a} that samples the Q-space in a spherical shell. These approaches, as well as more complex multi-shell sampling strategies, aim to construct and sharpen an orientation distribution function (ODF) in order to provide better approximate of the underlying diffusion. These methods require complex and long DWI acquisitions that are often unsuitable for clinical applications. Alternatively, methods based on spherical deconvolution (SD) can be used on HARDI scans of lesser angular resolution. The approach is to model the HARDI signal as a convolution of the fibre orientations. The resulting deconvoluted fibre orientation distribution (FOD) shows better discrimination of fibre directions compared to Q-ball under similar scanning parameters, but is susceptible to image noise\cite{Anderson2005,DellAcqua2007,Tournier2004}. Alternatively, multi-tensor model-based approaches estimate the fibre configuration with the assumption that there are no more than two or three crossings in a given voxel. These include eXtended Streamline Tractography (XST), which can delineate lateral projections of the corticospinal tract in the motor cortex, in DWI scans with 50 gradient directions \cite{Qazi2009}. Other tractography algorithm include stochastically resolving tract propagation based on established white-matter prior probability (probabilistic tractography) \cite{Behrens2007}. Probabilistic approaches however are predominantly image-based, and result in a visitation volume image that needs to be visualized volumetrically.

@@

Finally, we deploy our system to the evaluation of various tractography methods in their ability to delineate neuroanatomy that are highly specific but difficult to image. These include supra and infratentorial structures such as: small white matter bundles in heavy crossing fibres (cranial nerves), pontine decussation (rubrocerebellar pathway) and curved central pathways (fornix, visual and auditory radiations). We recruited 5 neuroanatomists to evaluate and rate the resulting visual reports generated by SAGIT. We then compared the results of NOS scores and the human raters to assess the automated NOS score as a tractography reproducibility metric.

@@

The individual tractography models were also converted to binary spatial images, and the resulting images were transformed to the template space. Multiple binary images were stacked together to form the conjunction percentage overlap image. The overlaps were then visualized using concentric isosurfaces created from step-wise (10\%) thresholds of the underlying overlap volume using the MayaVI data visualization library \cite{Ramachandran2011}. A colour lookup table with even visual brightness falloff \cite{VanderWalt2015} was chosen to avoid visual judgment biases from the colour presentation. Different viewpoints (axial, coronal, sagittal, and perspective) were created for each visualization, and then composed together to create the visual report.

@@

For the CN VII/VIII, CST$_{prob}$, CST$_{det}$ and XST were all able to delineate the cranial as well as brainstem segments of the nerve (Figure 2). CST$_{prob}$ however also delineated much of the cerebellum. DTT by Slicer was not able to delineate the brainstem portions of the fibre and therefore was scored lower than the other methods. These differences in delineation can be more clearly seen in the merged tracts (Figure 3), where the XST also showed further lateral delineations of CN VII/VIII. In unfiltered tractography, there was much more inter-rater variability for all except CST$_{prob}$. CST$_{det}$ and XST also showed lower average rating when unfiltered (Figure 5). The NOS showed similar trend where DTT resulted in the lowest score (Figure 6). The NOS also showed higher score for CST$_{det}$ when comparing to XST. There were no notable changes in NOS scores in unfiltered tractography.

@@

\begin{figure}[p]
\centering
\includegraphics[width=\linewidth]{figure2.png}
\caption[Example of the auto-generated visual panels.]{Example of the auto-generated visual panels. The panels that best represent the resulting anatomy are compiled. For anatomical reference, the best representative anatomy image slice is composed with the tract conjunction map. The colour scale of the conjunction image represents the percentage reproducible for each region, where 1=100\%. For each algorithm (rows), different visual perspectives are created. Together 6 different anatomies were delineated (columns). To see the full visual reports generated, please see supplemental.}
\label{fig:SGfigure2}
\end{figure}

@@

\afterpage{%
    \clearpage% Flush earlier floats (otherwise order might not be correct)
    \thispagestyle{empty}% empty page style (?)
    \begin{landscape}% Landscape page
\begin{figure}[p]
\includegraphics[width=\linewidth]{figure3.png}
\caption[Example of merged tractography models for each of the techniques and anatomy.]{Example of merged tractography models for each of the techniques and anatomy. The fornix is coloured to distinguish tracts from each of the 42 subjects, in order to demonstrate the effect of combining the group data. The other anatomies are coloured by FA intensity. The specific FA measures from each point were sampled from their native DWI space before tractography deformation. The grouping of the similar FA measures across the subjects show that the registrations were accurate. The Red Nuclei shows different metrics (FA and AD) depending on lateralization, in order to highlight the pontine decussation.}
\centering
\label{fig:SGfigure3}
\end{figure}
    \end{landscape}
    \clearpage% Flush page
}

@@

It can be observed that results of averaged tractography image conjunction across a group are visually similar to probabilistic tractography in an individual, and that the group junction image seems to stabilize as the number of subject increase. It can be argued that the conjunction image captures the probable volume across a sampled population for a particular neuroanatomy in the template space. Their exact similarity and differences from probabilistic tractography at a group level should be explored in future studies. Additionally, the fidelity of the merged anatomy is often higher than what would be available from an individual. The result supports the validity of this technical approach. It is possible that the resulting average delineations can be used to generate population-specific anatomy atlases. It's also possible to use fibre density images rather than binary mask for conjunction. However the result of an average fibre density image is difficult to interpret, and therefore should be explored in future studies.

@@

For the specific tractography algorithms, CST$_{prob}$ produces the highest ratings, and performs particularly well in the optical radiation. It also results in more false-positives. This is most evident in CN VII/VIII and CN X results. In these cases the cranial nerve anatomies are very specific and local, and yet CST$_{prob}$ produces more false-positive projections to distant regions. In the fornix there are also erroneous tract extensions into the corpus callosum. Deterministic methods in comparison are more conservative and therefore result in lower scores. They perform well when the anatomy is regional and has finer features. Both CST$_{det}$ and XST performed well with cranial nerves. XST is able to image the rubro-cerebellar pontine decussation with more reliability than CST$_{det}$. While CST$_{prob}$ decussation delineation shows the highest reproducibility, it also produces wider projection coverage that’s harder to interpret. The low DTT score is not surprising, as the single-tensor's inability to resolve crossing-fibres is well known. Based on the result of the study, we recommend the use of CST$_{prob}$ for tasks in which the target anatomy is well defined and false positively can be easily recognized , or when interhemispheric and long projection distances are desired. For exploratory tractography, where filtering locations are not well defined, or when the anatomy features are close together, deterministic methods are recommended. When comparing CST$_{det}$ and XST, it appears that XST is more conservative and thereby results in less false-positives. It also has better ability in resolving hemispherical decussations, as evident by the results of the rubro-cerebellar projections. In practice the two methods are very close in performance. Note however that the comparison is not exhaustive of the possible parameter combinations on either method, and therefore automated tuning of the parameters can answer this question more definitively.

### 2018-04-26 contents/tn_sagit.tex
Updates from ShareLaTeX

The trigeminal sensory system involves both discriminative touch and pain pathways \cite{Henssen2016}.  Pain and discriminitive touch stimuli trigger innervations of the three trigeminal nerve branches (V1, V2, and V3) in the head and face, these primary affrents converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion) in the middle cranial fossa. The discriminative touch afferents then course through the cisternal segment of CN V, enter the pons, and synapse onto the primary trigeminal sensory nucleus, after which they project to the contralateral VC thalamus and the S1 cortex. The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, after which it courses towards the fibers of the cingulate cortex.

The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations. However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status \cite{Hung2017} and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain \cite{Miller2009}. It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.

@@

\subsubsection{Tractography}
ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (specifications: ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibers must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary and exclude the brainstem).

@@

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. We have previously created a software library\footnote{https://github.com/sinkpoint/fascicle} to facilitate this process. Detailed steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group associations preserved. Finally, all template-space tractography models across all the subjects are merged into a single file. 
The merged tract was post-processed in 3D Slicer to remove anatomically erroneous streamlines. These include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise. TPT streamlines that did not decussate at the level of the pons, as well as fibers that stray into the contralateral TGN were removed. S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{cnv-inferior-view-annotated.png}
\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A). The expected increase in FA as the fibers enter the brainstem can be visualized, and denotes good registration between subjects. }
\label{fig:GPfigure1}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{view-left-right.png}
\caption[Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.]{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex. Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A). FA is depicted by the colour intensities (Red: right side; Blue: left side). The trigeminal nerves, including the cisternal segments are seen in a coronal perspective (CN V), following which two pathways can be seen. The first pathway appears to head into the spine, while the second pathway consists of the TPT. Following the decussation, the fibers proceed to the area of the sensory thalamus and from there to the cortical sensory area S1.}
\label{fig:GPfigure3}
\end{figure}

\subsubsection{Diffusivity Differences}
\paragraph{CN V}
The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves. Both sides were distinguished from controls with classifiers that achieved near 80\% accuracy (Figure \ref{fig:GPfigure4}). 
Classifiers on the right (affected) side had accuracies ranging from 76\% (GFA) to 86\% (FA). The left (unaffected) classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA). The right side showed greater overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substance of the trigeminal fibers in the brainstem. Overall, the windows of the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem. FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment. The GFA classifier achieved 70\% accuracy in region 9--12. Demonstrating that affected-unaffected diffusivity differences were found primarily in the cistern/REZ (Table \ref{table:GP}).

@@

The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (\textgreater 90\%), with overlapping windows in the ascending segment. The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric. The asymmetry of this decussation has not been demonstrated before. It is not clear whether this represents the anatomical arrangement of these fibers, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards). Further assessment of other decussation points will be helpful to understand this process.

We were able to successfully delineate the course of the fibers from the thalamus to S1 given the use of SAGIT and incorporation of Freesurfer cortical segmentation maps. Previous studies have suggested that S1 is involvement in pain location discrimination \cite{bushnell1999pain}, and have been found to exhibit grey-matter abnormalities in TN pain \cite{Desouza2013c}. We found that that while the left (affected) S1 RD classifier had higher accuracy (87\%) compare to the right S1 RD classifier (83\%), the left-verus-right classifier windows showed more overlap in subdivisions close to the thalamus. This suggests that while there is bilateral difference in the S1 pathways from controls, the difference between the segments mitigates as these reach closer to the S1 cortex.

@@

FA classifiers consistently achieved the highest accuracy in CNV and TPT, with the exception of S1. This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibers. However, the TPT pathway also situations in regions of complex crossing fibers. One explanation for FA classifier in TPT is that the pathway courses through the narrow passage of pontine/thalamic regions, the fiber structures are highly constrained. Therefore FA still remains as a sensitive diffusivity metric for the TPT.

We included GFA to investigate its potential for measuring cross-fiber white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric. Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fiber regional dynamics. For example, in the crossing-fiber region of healthy tissue, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher. In a pathological cross region, where only one bundle is healthy, the FA increases due to the diffusivity bias in one direction; the RD will be lower, and MD will be lower as well. Moreover, in diverging and converging structures, as well as in multi-crossing regions such as the corona radiata, the interpretations of Gaussian-based diffusivity metrics become even more unclear. However results demonstrated that GFA did not perform to any substantially degree than the Gaussian-based diffusivity metrics such as FA.

@@

\subsubsection{Limitations}
The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord. The consistent observation of diverging streamlines at the levels of the medulla strongly suggests that these fibers correspond to trigeminal spinal decussations. However the caudal projection were not consistently seen. This is likely due to the limited size of voxel resolution that in turn does not permit adequate resolution of medullary and spinal cord fiber details.

>>>>

The trigeminal sensory system involves both discriminative touch and pain pathways \cite{Henssen2016}.  Pain and discriminitive touch stimuli trigger innervations of the three trigeminal nerve branches (V1, V2, and V3) in the head and face, these primary affrents converge and form the trigeminal root gangion (TRG) (also known as Gasserian, or semilunar ganglion) in the middle cranial fossa. The discriminative touch afferents then course through the cisternal segment of CN V, enter the pons, and synapse onto the primary trigeminal sensory nucleus, after which they project to the contralateral VC thalamus and the S1 cortex. The nociceptive pathway synapses caudally onto the trigeminal spinal nucleus, decussates at the level of the medulla and spinal cord. The pathway merges with the discriminative sensory pathway and ascend as the contralateral trigeminal lemniscus, and also project to the contralateral VC thalamus. The pain pathway also projects to the medial nucleus of the thalamus, after which it courses towards the fibres of the cingulate cortex.

The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations. However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status \cite{Hung2017} and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain \cite{Miller2009}. It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibres.

@@

\subsubsection{Tractography}
ROIs were defined on the template image to delineate the following bilateral structures: 1) CN V (specifications: ROI placed at cistern nerve root; excludes regions of cerebellum grey matter and pontine cerebellar peduncles); 2) trigeminothalamic pontine decussation (TPT) (ROIs placed at trigeminal nucleus (TGN); the fibres must traverse pontine decussation towards contralateral VC thalamus; pathway must exclude ipsilateral thalamus, cerebellar grey matter, pons, and contra-lateral S1 ); 3) thalamocortical S1 pathway (ROIs placed at VC thalamus; pathway must traverse the ipsilateral S1 white-matter boundary and exclude the brainstem).

@@

\subsubsection{Merged Tractography}
All tractography delineations of the same anatomy across all the subjects and groups were merged into a single tractography file for analysis. We have previously created a software library\footnote{https://github.com/sinkpoint/fascicle} to facilitate this process. Detailed steps are as follows: The individual tract files were converted into a SQLite relational database \cite{owens2010sqlite}, and  all points were exported and transformed by applying the subject DWI to T1, and T1 to template transforms using ANTS' applyTransformToPoints command \cite{Avants2009}. The transformed points were then imported back into the  database, and a new template-space tractography model is exported using the transformed point-set, with all its original scalars and subject/group associations preserved. Finally, all template-space tractography models across all the subjects are merged into a single file. 
The merged tract was post-processed in 3D Slicer to remove anatomically erroneous streamlines. These include remnant CN V streamlines that stray into the cerebellar peduncles or other non-TGN regions, and streamlines that are perpendicular to the nerve in the cisternal space that appear to be partial delineation of vessels or noise. TPT streamlines that did not decussate at the level of the pons, as well as fibres that stray into the contralateral TGN were removed. S1 streamlines that stray medially towards other regions such as the corpus callosum were also removed.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{cnv-inferior-view-annotated.png}
\caption{Merged CN V reconstruction of 36 subjects, with the colour scale depicting FA values derived from raw streamlines, prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A). The expected increase in FA as the fibres enter the brainstem can be visualized, and denotes good registration between subjects. }
\label{fig:GPfigure1}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{view-left-right.png}
\caption[Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex.]{Visual representation of the course of the trigeminal pathway from its cisternal component to the face area of the sensory S1 cortex. Images obtained prior to the generation of centroids (see Figure \ref{fig:GPfigure-method-centroid} A). FA is depicted by the colour intensities (Red: right side; Blue: left side). The trigeminal nerves, including the cisternal segments are seen in a coronal perspective (CN V), following which two pathways can be seen. The first pathway appears to head into the spine, while the second pathway consists of the TPT. Following the decussation, the fibres proceed to the area of the sensory thalamus and from there to the cortical sensory area S1.}
\label{fig:GPfigure3}
\end{figure}

\subsubsection{Diffusivity Differences}
\paragraph{CN V}
The classifier distinguished bilateral trigeminal nerves (affected and unaffected) from controls as well as affected trigeminal nerve from unaffected nerves. Both sides were distinguished from controls with classifiers that achieved near 80\% accuracy (Figure \ref{fig:GPfigure4}). 
Classifiers on the right (affected) side had accuracies ranging from 76\% (GFA) to 86\% (FA). The left (unaffected) classifiers achieved accuracy ranging from 79\% (MD) to 81\% (FA). The right side showed greater overlap of segment ranges between positions 14--24, which corresponds to the region of the cistern/ganglion; while the left side showed segments that covered the entire the nerve, and extended to cover position 0--4, within the substance of the trigeminal fibres in the brainstem. Overall, the windows of the right side classifiers extend further into the distal subdivisions, into the cistern and the area of the ganglion, while the left side extended more into the brainstem. FA classifier consistently achieved the highest accuracy in both sides, with right side at position 17--23 (cistern/ganglion) and left FA at 2--16 (brainstem/cistern).
The FA classifier achieved 75\% accuracy in differentiating affected versus unaffected nerves, where its segments ranged from 11 to 19, which was the cisternal segment. The GFA classifier achieved 70\% accuracy in region 9--12. Demonstrating that affected-unaffected diffusivity differences were found primarily in the cistern/REZ (Table \ref{table:GP}).

@@

The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (\textgreater 90\%), with overlapping windows in the ascending segment. The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric. The asymmetry of this decussation has not been demonstrated before. It is not clear whether this represents the anatomical arrangement of these fibres, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards). Further assessment of other decussation points will be helpful to understand this process.

We were able to successfully delineate the course of the fibres from the thalamus to S1 given the use of SAGIT and incorporation of Freesurfer cortical segmentation maps. Previous studies have suggested that S1 is involvement in pain location discrimination \cite{bushnell1999pain}, and have been found to exhibit grey-matter abnormalities in TN pain \cite{Desouza2013c}. We found that that while the left (affected) S1 RD classifier had higher accuracy (87\%) compare to the right S1 RD classifier (83\%), the left-verus-right classifier windows showed more overlap in subdivisions close to the thalamus. This suggests that while there is bilateral difference in the S1 pathways from controls, the difference between the segments mitigates as these reach closer to the S1 cortex.

@@

FA classifiers consistently achieved the highest accuracy in CNV and TPT, with the exception of S1. This might be due to the fact that it is known to have trouble in regions of heavy crossing-fibres. However, the TPT pathway also situations in regions of complex crossing fibres. One explanation for FA classifier in TPT is that the pathway courses through the narrow passage of pontine/thalamic regions, the fibre structures are highly constrained. Therefore FA still remains as a sensitive diffusivity metric for the TPT.

We included GFA to investigate its potential for measuring cross-fibre white-matter diffusivity, as it is the most well-known non-Gaussian diffusivity metric. Although metrics such as FA are popular in current neuroimaging literature, they nevertheless are derived from the Gaussian tensor model, which cannot adequately describe cross-fibre regional dynamics. For example, in the crossing-fibre region of healthy tissue, where both bundles have equal myelin and axonal integrity, will have a low FA score due to a spherical tensor; its AD and RD will be roughly equal, and therefore MD will be higher. In a pathological cross region, where only one bundle is healthy, the FA increases due to the diffusivity bias in one direction; the RD will be lower, and MD will be lower as well. Moreover, in diverging and converging structures, as well as in multi-crossing regions such as the corona radiata, the interpretations of Gaussian-based diffusivity metrics become even more unclear. However results demonstrated that GFA did not perform to any substantially degree than the Gaussian-based diffusivity metrics such as FA.

@@

\subsubsection{Limitations}
The merged CN V delineations of all the subjects showed caudal projections towards the spinal cord. The consistent observation of diverging streamlines at the levels of the medulla strongly suggests that these fibres correspond to trigeminal spinal decussations. However the caudal projection were not consistently seen. This is likely due to the limited size of voxel resolution that in turn does not permit adequate resolution of medullary and spinal cord fibre details.
